DBH:Height models for EP1162

### 2020-03-27 analysis/heightModel.Rmd
volume estimates

---
title: "Height estimates from DBH:Height relationship"
author: "Hardy Griesbauer"
date: "20/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

# Load data
load(here("data","ep1162_Data.RData"))

@@

1. Model DBH:Height relationships separately for each measurement period, and for Bl:Sx separately (at first).
2. Compare Bl and Sx relationships, compare estimates between years, and determine if we can pool species and/or years to estimate Height.

@@

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) %>% # remove trees where height was not measured

@@

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
    drop_na(Height) %>% # remove trees where height was not measured

@@

# Assign data
x<-
   dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
  filter(Year==year) %>% 
    drop_na(Height) # remove trees where height was not measured

@@

rbind(speciesMod1(1992)$out,
      speciesMod1(1994)$out,
      speciesMod1(1997)$out,
      speciesMod1(2019)$out) %>% 
  as_tibble()

@@

Here are the coefficients from Mike's analysis:

*Subalpine fir*
HT = 11.188 x ln(DBH)  - 17.609

*Hybrid white spruce*
HT = 11.79 x ln(DBH)  - 19.637

>>>>

---
title: "DBH:Height models for EP1162"
author: "Hardy Griesbauer"
date: "20/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)

@@

# Load data
load(here::here("data","ep1162_Data.RData"))

@@

1. Model DBH:Height relationships separately for each measurement period, and for Bl:Sx separately (at first).
2. Compare Bl and Sx relationships, and determine if we can pool species to estimate Height.

We will generate different dbh:height models for each measurement year, as pooling years would violate assumptions of independence.

### Data preparation: remove trees with broken tops
Trees with broken or dead tops that have height measurements should not be included in DBH:height models.  We can summarise and remove these:

```{r}
dat %>% 
  filter(!is.na(DBT)) %>% # filter for trees with recorded broken top
  filter(!is.na(Height)) %>% # filter for trees with height measurements
  group_by(Species,Year) %>% # summarise trees with broken tops at the species and year level
  summarise(Num.DBT=n()) %>% 
  pivot_wider(names_from="Species",values_from="Num.DBT")

@@

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured

@@

  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
   filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured

@@

# Assign data
x<-
   dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% 
   filter(is.na(DBT)) %>% # filter for trees with no broken tops
  filter(Year==year) %>% 
    drop_na(Height) # remove trees where height was not measured

@@

heightModels<- # assign 
  rbind(speciesMod1(1992)$out,
      speciesMod1(1994)$out,
      speciesMod1(1997)$out,
      speciesMod1(2019)$out) %>% 
  as_tibble()

print(heightModels)

@@

#### Height model for 2009 measurement period
Height was not recorded during the 2009 measurements, however, we should be able to estimate heights using the 2009 DBH data.  I propose the following model to estimate 2009 heights:

### Graph out intercept and slope for four measurement period
I noticed that the intercept and slope vary monotonically with years, which suggests that we might be able to predict intercept/slope for a given year (in this case 2009).  Let's examine this with a scatterplot:

```{r}

heightModels %>% 
  dplyr::select(1:3) %>% #select first three columns
  pivot_longer(-year,values_to="coefficient",names_to="component") %>% # pivot longer for facet plot
  ggplot()+
  aes(x=year,y=coefficient)+
  geom_point()+
  geom_line(color="red")+
  geom_vline(xintercept=2009,linetype="dashed")+
  facet_wrap(~component,nrow=2,scales="free")

Based on these relationships, it seems we can reasonably estimate the slope and intercept for a dbh:height model at year 2009.  Thoughts?

### Calculate 2009 intercept and slope coefficients for dbh:height model

```{r}

# Calculate slope to estimate intercept:
int.slope<-
  with(heightModels,heightModels[year==2019,"intercept"]-heightModels[year==1997,"intercept"])/(2019-1997)
  int.slope$run=diff(c(1997,2009))
  int.2009<-  as.numeric(with(int.slope,run*intercept)+with(heightModels,heightModels[year==1997,"intercept"]))

####
slope<-
  with(heightModels,heightModels[year==2019,"slope"]-heightModels[year==1997,"slope"])/(2019-1997)
  slope$run=diff(c(1997,2009))
  slope.2009<-  as.numeric(with(slope,run*slope)+with(heightModels,heightModels[year==1997,"slope"]))

### Add these values into heightModels table
  heightModels<-
    heightModels %>% 
    add_row(year=2009,intercept=int.2009,slope=slope.2009) %>% # add 2009 coefficients to table
    arrange(year) # order by year

# print to screen
  heightModels

### Save file for volume estimates
```{r}
# save coefficients for volume estimates
save(heightModels,file=here("data","heightModels.RData"))
```
